---
title: "Exploratory Data Analysis (EDA)"
description: |
  A description and graphical exploration of the datasets used in the research
site: distill::distill_website
output:
  distill::distill_article:
    code_folding: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(visdat)
library(corrplot)
library(tidyr)
library(rmarkdown)
library(report)


precipitation <- 
  read_csv("data/hardvard_dataverse_tropidry/Daily_Precipitation_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd(date),
         do_y = as.numeric(do_y),
         precip = as.numeric(precip))

biomet <- 
  read_csv("data/hardvard_dataverse_tropidry/Micrometeorological_Biomet_Parameters_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd_hm(date)) %>% 
  mutate(across(where(is.character), as.numeric))

monthly_gpp <- 
  read_csv("data/hardvard_dataverse_tropidry/Monthly_NEE_GPP_Reco_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  unite(col = date, year, month, sep = "-") %>% 
  mutate(date = paste0(date, "-1"),
         date = ymd(date)) %>% 
  select(-date_1, -date_2) %>% 
  mutate(across(where(is.character), as.numeric))

ndvi_moha <- 
  readxl::read_xlsx("data/hardvard_dataverse_tropidry/final -NDVI- smoothed -complete.xlsx") %>% 
  clean_names()

modis_gpp <- 
  read_csv("data/modis-gpp-sr/modis-sr-MYD17A2H-006-results.csv") %>% 
  clean_names() %>% 
  select(date, myd17a2h_006_gpp_500m, 
         myd17a2h_006_psn_qc_500m_modland_description,
         myd17a2h_006_psn_qc_500m_cloud_state_description,
         myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  rename(gpp = myd17a2h_006_gpp_500m,
         modland_description = myd17a2h_006_psn_qc_500m_modland_description,
         cloud_state_description = myd17a2h_006_psn_qc_500m_cloud_state_description,
         scf_qc_description = myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  mutate(gpp = gpp * 10)

modis_indices <- 
  read_csv("data/modis-indices-sr/modis-indices-sr-MOD13Q1-006-results.csv") %>% 
  clean_names() %>% 
  select(date, mod13q1_006_250m_16_days_evi, 
         mod13q1_006_250m_16_days_ndvi, 
         mod13q1_006_250m_16_days_vi_quality,
         mod13q1_006_250m_16_days_pixel_reliability,
         mod13q1_006_250m_16_days_pixel_reliability_modland_description) %>% 
  rename(evi = mod13q1_006_250m_16_days_evi,
         ndvi = mod13q1_006_250m_16_days_ndvi,
         vi_quality = mod13q1_006_250m_16_days_vi_quality,
         pixel_reliability = mod13q1_006_250m_16_days_pixel_reliability,
         reliability_modland_description =
           mod13q1_006_250m_16_days_pixel_reliability_modland_description)
```

## SRNP data exploration

Santa Rosa National Park Environmental Super Site in Guanacaste, Costa Rica
have several sensors to monitor environmental climate variables, radiation and
fluxes.

Given that variables such as NDVI and GPP can be calculate with instrumentation
in the site, this ones will be included and name as `in-situ_variable_name` to
differentiate them from the same variables obtained from the MODIS satellite
images.

| Dataset | Frequency | Date range |
| ------- | --------- | ---------- |
| Precipitation | Daily values | `2013-05-15 00:00:00 UTC` to `2017-05-17 23:30:00 UTC`|
| Meteorological data | Half hour values | `2013-05-15 00:00:00 UTC` to `2017-05-17 23:30:00 UTC` |
| Monthly Gross Primary Production (GPP) | Monthly values | `2013-05-01` to `2017-05-01`|
| NDVI (Normalized Difference Vegetation Index) | Daily values | `2013-02-23 UTC` to `2022-01-01 UTC`

### Santa Rosa Data tables glimpse

For Santa Rosa we have the following data sets:

#### Precipitation data

```{r}
precipitation %>% 
  slice(1:100) %>% 
  paged_table()
```

#### Meteorological data

```{r}
biomet %>% 
  slice(1:100) %>% 
  paged_table()
```

#### GPP in-situ data

```{r}
monthly_gpp %>% 
  slice(1:100) %>% 
  paged_table()
```

#### NDVI in-situ data

```{r}
ndvi_moha %>% 
  slice(1:100) %>% 
  paged_table()
```

### Precipitation EDA

```{r fig.cap = "Total precipitation (mm) for Santa Rosa National Park. Year 2015 had the lowest values of precicipitation"}
precipitation %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  group_by(year_mon) %>% 
  summarize(
    total_precip = sum(precip, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = as.factor(year_mon), y = total_precip)) +
  geom_bar(stat = "identity") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "Total precipitation per month (mm)")
```


```{r fig.cap = "Total precipitation (mm) arrange per month for comparison per each year. Patterns for 2014 and 2015 as the months with less precipitation are more evident"}
### Agrupando los meses
precipitation %>%
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  filter(year(date) < 2017) %>% 
  group_by(year_mon) %>%
  summarize(
    total_precip = sum(precip, na.rm = TRUE)
  ) %>%
  mutate(year = year(year_mon),
         month = month(year_mon, label = TRUE)) %>%
  # mutate(year_mon = as.character(year_mon)) %>%
  ggplot(aes(x = as.factor(month), y = total_precip, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_light(base_size = 12) +
  scale_y_continuous(breaks = seq(0, 450, by = 50)) +
  scale_fill_viridis_d() +
  labs(x = "Date", y = "Total precipitation per month (mm)",
       fill = "Year")
```

### Environmental drivers

```{r fig.cap = "Seasonal pattern of VPD (hPa). Months asociated with lower precipitation rates have the highest values of VPD"}
biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  filter(vpd > 0) %>% 
  ggplot(aes(x = as.factor(year_mon), y = vpd, fill = as.factor(year_mon))) +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(-10, 40, by = 5)) +
  scale_fill_viridis_d() +
  theme_light(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "Vapor presurre deficit (hPa)")
```


```{r}
biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  # filter(vpd > 0) %>% 
  ggplot(aes(x = as.factor(year_mon), y = tair, fill = as.factor(year_mon))) +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(15, 38, by = 2)) +
  scale_fill_viridis_d() +
  theme_light(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "Air temperature (C)")
```


```{r}
biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  filter(le > 0) %>%
  ggplot(aes(x = as.factor(year_mon), y = le, fill = as.factor(year_mon))) +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  # scale_y_continuous(breaks = seq(15, 38, by = 2)) +
  scale_fill_viridis_d() +
  theme_light(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "Latent Heat (Wm-2)")
```


```{r}
biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  # filter(le > 0) %>%
  ggplot(aes(x = as.factor(year_mon), y = r_h, fill = as.factor(year_mon))) +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  # scale_y_continuous(breaks = seq(15, 38, by = 2)) +
  scale_fill_viridis_d() +
  theme_light(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "Relative Humidity (%)")
```

### NDVI

```{r}
ndvi_moha %>% 
  mutate(date = ymd(date)) %>% 
  filter(year(date) < 2017) %>% 
  ggplot(aes(x = date, y = ndvi)) +
  # geom_point(alpha = 0.5, color = "#FF3A1D", size = 3) +
  geom_jitter(alpha = 0.5, color = "#FF3A1D", size = 2.5) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1))
```

### Monthly GPP

```{r}
monthly_gpp %>% 
  ggplot(aes(x = date, y = average_gpp, group = 1)) +
  geom_errorbar(aes(ymin = average_gpp - gpp_stdev,
                    ymax = average_gpp + gpp_stdev),
                colour = "#4D4D4D", width = .4) +
  geom_line(linetype = 2) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "GPP (mgm-2s-1)")
```

### Relation between in-situ NDVI and in-situ GPP

```{r}
ndvi_summarized <- ndvi_moha %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  group_by(year_mon) %>% 
  summarize(
    ndvi_mean = mean(ndvi, na.rm = TRUE),
    ndvi_sd = sd(ndvi, na.rm = TRUE)
  )

gpp_ndvi <- monthly_gpp %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  inner_join(ndvi_summarized) 

gpp_ndvi %>% 
  ggplot(aes(x = average_gpp, y = ndvi_mean)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  theme_linedraw() +
  geom_smooth(method = "lm")


model <- lm(average_gpp ~ ndvi_mean, data = gpp_ndvi)
# summary(model)
report(model)
```

<!-- Puedo sacar la regresión lineal entre gpp y ndvi y con esta sacar los -->
<!-- distintos valores de gpp para cada día por ejemplo. -->
<!-- http://www.sthda.com/english/articles/40-regression-analysis/165-linear-regression-essentials-in-r/ -->

## MODIS data exploration

MODIS data collected for the site have 3 variables of interest:

 - NDVI
 - EVI
 - GPP

### MODIS Data tables glimpse

Data derived from satellite images for Santa Rosa is compose of following data
sets:

#### MODIS indices data

```{r}
modis_indices %>% 
  slice(1:100) %>% 
  paged_table()
```

#### MODIS GPP data

```{r}
modis_gpp %>% 
  slice(1:100) %>% 
  paged_table()
```

### GPP EDA

```{r}
modis_gpp %>% 
  ggplot(aes(x = date, y = gpp, group = 1)) +
  geom_line(linetype = 2) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "GPP (mgm-2s-1)")
```

### NDVI EDA

```{r}
modis_indices %>% 
  filter(reliability_modland_description !=
           "Marginal data, Useful, but look at other QA information") %>% 
  mutate(date = ymd(date)) %>% 
  filter(ndvi > 0.2) %>% 
  ggplot(aes(x = date, y = ndvi)) +
  geom_jitter(alpha = 0.7, color = "#FF3A1D", size = 3) +
  geom_line(linetype = 2) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "NDVI")
```


