---
title: "Exploratory Data Analysis (EDA)"
description: |
  A description and graphical exploration of the datasets used in the research
site: distill::distill_website
output:
  distill::distill_article:
    code_folding: true
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(visdat)
library(corrplot)
library(tidyr)
library(rmarkdown)
library(report)
library(cowplot)


precipitation <- 
  read_csv("data/hardvard_dataverse_tropidry/Daily_Precipitation_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd(date),
         do_y = as.numeric(do_y),
         precip = as.numeric(precip))

biomet <- 
  read_csv("data/hardvard_dataverse_tropidry/Micrometeorological_Biomet_Parameters_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd_hm(date)) %>% 
  mutate(across(where(is.character), as.numeric))

monthly_gpp <- 
  read_csv("data/hardvard_dataverse_tropidry/Monthly_NEE_GPP_Reco_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  unite(col = date, year, month, sep = "-") %>% 
  mutate(date = paste0(date, "-1"),
         date = ymd(date)) %>% 
  select(-date_1, -date_2) %>% 
  mutate(across(where(is.character), as.numeric))

ndvi_moha <- 
  readxl::read_xlsx("data/hardvard_dataverse_tropidry/final -NDVI- smoothed -complete.xlsx") %>% 
  clean_names()

modis_gpp <- 
  read_csv("data/modis-gpp-sr/modis-sr-MYD17A2H-006-results.csv") %>% 
  clean_names() %>% 
  select(date, myd17a2h_006_gpp_500m, 
         myd17a2h_006_psn_qc_500m_modland_description,
         myd17a2h_006_psn_qc_500m_cloud_state_description,
         myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  rename(gpp = myd17a2h_006_gpp_500m,
         modland_description = myd17a2h_006_psn_qc_500m_modland_description,
         cloud_state_description = myd17a2h_006_psn_qc_500m_cloud_state_description,
         scf_qc_description = myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  mutate(gpp = gpp * 10)

modis_indices <- 
  read_csv("data/modis-indices-sr/modis-indices-sr-MOD13Q1-006-results.csv") %>% 
  clean_names() %>% 
  select(date, mod13q1_006_250m_16_days_evi, 
         mod13q1_006_250m_16_days_ndvi, 
         mod13q1_006_250m_16_days_vi_quality,
         mod13q1_006_250m_16_days_pixel_reliability,
         mod13q1_006_250m_16_days_pixel_reliability_modland_description) %>% 
  rename(evi = mod13q1_006_250m_16_days_evi,
         ndvi = mod13q1_006_250m_16_days_ndvi,
         vi_quality = mod13q1_006_250m_16_days_vi_quality,
         pixel_reliability = mod13q1_006_250m_16_days_pixel_reliability,
         reliability_modland_description =
           mod13q1_006_250m_16_days_pixel_reliability_modland_description)
```

The **objective** of the Exploratory Data Analysis (EDA) is to check all the
data sets available, evaluate if there is a need to clean and process the data
further to be able to model and evaluate if there are patterns that I need to
take into account.

This EDA will be divided into two main sections: Santa Rosa National Park data
and MODIS derived data.

## SRNP data exploration

Santa Rosa National Park Environmental Super Site in Guanacaste, Costa Rica have
several sensors to monitor environmental climate variables and fluxes.

Given that variables such as NDVI and GPP can be calculate with instrumentation
in the site, this ones will be included and name as `in-situ_variable_name` to
differentiate them from the same variables obtained from the MODIS satellite
images.

The following table shows the data sets used to conduct this research. There are
3 different data sets which contains variables of interest. Not all data sets
have the same time range or the same time frequency for each observation:

+--------------------+--------------------------+-----------------------------+
| Dataset            | Frequency                | Date range                  |
+====================+==========================+=============================+
| Precipitation      | Daily values             | `2013-05-15 00:00:00 UTC`   |
|                    |                          | to                          |
|                    |                          | `2017-05-17 23:30:00 UTC`   |
+--------------------+--------------------------+-----------------------------+
| Meteorological     | Half hour values         | `2013-05-15 00:00:00 UTC`   |
| data               |                          | to                          |
|                    |                          | `2017-05-17 23:30:00 UTC`   |
+--------------------+--------------------------+-----------------------------+
| Monthly Gross      | Monthly values           | `2013-05-01` to             |
| Primary Production |                          | `2017-05-01`                |
| (GPP)              |                          |                             |
+--------------------+--------------------------+-----------------------------+

### Santa Rosa Data tables glimpse

This section contains tables that shows the first 100 rows and all the variables
contained in each of the data sets shown in the table above:

#### Precipitation data

```{r}
precipitation %>% 
  slice(1:100) %>% 
  paged_table()
```

#### Meteorological data

```{r}
biomet %>% 
  slice(1:100) %>% 
  paged_table()
```

#### Monthly GPP in-situ data

```{r}
monthly_gpp %>% 
  slice(1:100) %>% 
  paged_table()
```

<!-- #### NDVI in-situ data -->

```{r, echo =  FALSE, eval = FALSE}
ndvi_moha %>% 
  slice(1:100) %>% 
  paged_table()
```

### Precipitation seasonal patterns

The total annual precipitation in Santa Rosa National Park is between 700 and 
2000 mm, with a dry season of 4 to 5 months where precipitation is less than 
100 mm per month or even 0 mm (Sánchez-Azofeifa __et al.__, 2005). For the time
period of this research, the site experimented a drought season in 2014 and 2015
(Castro *et al.* 2018)

```{r fig.cap = "Total precipitation (mm) for Santa Rosa National Park. Year 2015 had the lowest values of precicipitation"}
precipitation %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  group_by(year_mon) %>% 
  summarize(
    total_precip = sum(precip, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = as.factor(year_mon), y = total_precip)) +
  geom_bar(stat = "identity") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "Total precipitation per month (mm)")
```

```{r fig.cap = "Total precipitation (mm) arrange per month for comparison per each year. Patterns for 2014 and 2015 as the months with less precipitation are more evident", echo = FALSE, eval = FALSE}
### Agrupando los meses
precipitation %>%
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  filter(year(date) < 2017) %>% 
  group_by(year_mon) %>%
  summarize(
    total_precip = sum(precip, na.rm = TRUE)
  ) %>%
  mutate(year = year(year_mon),
         month = month(year_mon, label = TRUE)) %>%
  # mutate(year_mon = as.character(year_mon)) %>%
  ggplot(aes(x = as.factor(month), y = total_precip, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_light(base_size = 12) +
  scale_y_continuous(breaks = seq(0, 450, by = 50)) +
  scale_fill_viridis_d() +
  labs(x = "Date", y = "Total precipitation per month (mm)",
       fill = "Year")
```

### Metereological variables

From the site, there is instrumentation to measure environmental variables such
as Vapor Pressure Deficit (VPA), Air Temperature (TA), Latent Heat (LH), 
relative humidity (RH) and Soil Water Content (SWC). In order to explore these 
variables, I plotted them with box plots to check the variability outliers and 
patterns per month, across all the time range available (May 2013 to May 2017).

```{r fig.cap = "Seasonal pattern of environmental variables at Santa Rosa National Park. Yellow background denotes the year with less precipitation", layout="l-body-outset", fig.height = 9}

vpd_plot <- biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  filter(vpd > 0) %>% 
  ggplot(aes(x = as.factor(year_mon), y = vpd)) +
  geom_rect(aes(xmin = "Jan 2015",
                xmax = "Jan 2016",
                ymin = 0,
                ymax = Inf),
            alpha = 0.05,
            fill = "#D8E082") +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(-10, 40, by = 5)) +
  scale_fill_viridis_d() +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "VPD (hPa)")

at_plot <- biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  ggplot(aes(x = as.factor(year_mon), y = tair)) +
  geom_rect(aes(xmin = "Jan 2015",
                xmax = "Jan 2016",
                ymin = 15,
                ymax = Inf),
            alpha = 0.05,
            fill = "#D8E082") +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(15, 38, by = 2)) +
  scale_fill_viridis_d() +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "TA (C)")

lh_plot <- biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  filter(le > 0) %>%
  ggplot(aes(x = as.factor(year_mon), y = le)) +
  geom_rect(aes(xmin = "Jan 2015",
                xmax = "Jan 2016",
                ymin = 0,
                ymax = Inf),
            alpha = 0.05,
            fill = "#D8E082") +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "LH (Wm-2)")

rh_plot <- biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  ggplot(aes(x = as.factor(year_mon), y = r_h)) +
  geom_rect(aes(xmin = "Jan 2015",
                xmax = "Jan 2016",
                ymin = 0,
                ymax = Inf),
            alpha = 0.05,
            fill = "#D8E082") +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "RH (%)")

swc_plot <- biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>%
  filter(swc > 0) %>%
  ggplot(aes(x = as.factor(year_mon), y = swc)) +
  geom_rect(aes(xmin = "Jan 2015",
                xmax = "Jan 2016",
                ymin = 0,
                ymax = Inf),
            alpha = 0.05,
            fill = "#D8E082") +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  guides(fill = "none") +
  labs(x = "Date", y = "SWC (%)")


plot_grid(vpd_plot, at_plot, lh_plot, rh_plot, swc_plot,
          nrow = 5,
          labels = "AUTO",
          label_size = 10,
          align = "v"
)
```

<!-- ### NDVI -->

```{r, echo = FALSE, eval = FALSE}
ndvi_moha %>% 
  mutate(date = ymd(date)) %>% 
  filter(year(date) < 2017) %>% 
  ggplot(aes(x = date, y = ndvi)) +
  # geom_point(alpha = 0.5, color = "#FF3A1D", size = 3) +
  geom_jitter(alpha = 0.5, color = "#FF3A1D", size = 2.5) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1))
```

### Monthly GPP estimated in-situ

Monthly GPP values for the Santa Rosa National Park were obtained from the
Harvard dataverse, as a result of the publication from Castro *et al.* 2018 were
GPP was estimated using a light-response curve. Values presented here are the
mean per month.

```{r, fig.cap = "Monthly GPP values estimated in-situ for Santa Rosa National Park"}
monthly_gpp %>% 
  ggplot(aes(x = date, y = average_gpp, group = 1)) +
  geom_errorbar(aes(ymin = average_gpp - gpp_stdev,
                    ymax = average_gpp + gpp_stdev),
                colour = "#4D4D4D", width = 20, size = 0.5) +
  geom_line(linetype = 2) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "GPP (mg m-2 s-1)")
```

<!-- ### Relation between in-situ NDVI and in-situ GPP -->

```{r, echo = FALSE, eval = FALSE}
ndvi_summarized <- ndvi_moha %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  group_by(year_mon) %>% 
  summarize(
    ndvi_mean = mean(ndvi, na.rm = TRUE),
    ndvi_sd = sd(ndvi, na.rm = TRUE)
  )

gpp_ndvi <- monthly_gpp %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  inner_join(ndvi_summarized) 

gpp_ndvi %>% 
  ggplot(aes(x = average_gpp, y = ndvi_mean)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  theme_linedraw() +
  geom_smooth(method = "lm")


model <- lm(average_gpp ~ ndvi_mean, data = gpp_ndvi)
# summary(model)
# report(model)
```

<!-- Puedo sacar la regresión lineal entre gpp y ndvi y con esta sacar los -->

<!-- distintos valores de gpp para cada día por ejemplo. -->

<!-- http://www.sthda.com/english/articles/40-regression-analysis/165-linear-regression-essentials-in-r/ -->

## MODIS data exploration

MODIS data consist of 2 datasets: one with the GPP observations and a second one
with the indices (NDVI and EVI) observations. In this section I will explore the
data quality and possible relations between those variables.

### MODIS data tables glimpse

Data derived from satellite images for Santa Rosa is compose of following data
sets:

#### MODIS indices data

```{r}
modis_indices %>% 
  slice(1:100) %>% 
  paged_table()
```

#### MODIS GPP data

```{r}
modis_gpp %>% 
  slice(1:100) %>% 
  paged_table()
```

#### Observations per month

```{r}
## This takes from 94 observations to 53 observations. Remove 41 observations
modis_indices_clean <- modis_indices %>% 
  filter(reliability_modland_description == "Good data, use with confidence") %>% 
  select(date, ndvi, evi) 

## Filter observations with good quality
## This takes from 183 observations to 114. Remove 69 observations.
modis_gpp_clean <- modis_gpp %>% 
  filter(modland_description == "Good quality",
         cloud_state_description == "Significant clouds NOT present (clear)",
         scf_qc_description == "Very best possible") %>% 
  select(date, gpp) 
```

Given that satellite images from MODIS consist of at least 2 observations per
month for the site, we need to evaluate how many data points I have for the
analysis after filtering out observations with low quality (clouds present and
image quality assurance flags marked as bad)

From MODIS I have two datasets (one composed with GPP observations and a second
one composed with NDVI and EVI indices observations) the process was applied for
both datasets. In the case of GPP I had in total `r nrow(modis_gpp)` and after
filtering observations without good data quality the total observations
remaining are `r nrow(modis_gpp_clean)`.

For the EVI and NDVI data set, originally I had `r nrow(modis_indices)` and
after the filtering, I obtained `r nrow(modis_indices_clean)` in total for
further analysis.

```{r, fig.cap = "MODIS NDVI and EVI observations after filtering bad quality data points from both datasets"}
modis_indices_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  tally() %>% 
  rename("date" = `zoo::as.yearmon(date)`, "total" = "n") %>% 
  ggplot(aes(x = as.factor(date),
             y = total))  +
  geom_bar(stat = "identity", fill = "#75AADB") +
  scale_y_continuous(breaks = seq(0, 3, by = 1)) +
  labs(x = "Date",
       y = "Total  MODIS indices observations") +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1))
```


```{r, fig.cap = "MODIS GPP observations after filtering bad quality data points from both datasets"}
modis_gpp %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  tally() %>% 
  rename("date" = `zoo::as.yearmon(date)`, "total" = "n") %>% 
  ggplot(aes(x = as.factor(date),
             y = total))  +
  geom_bar(stat = "identity", fill = "#75AADB") +
  scale_y_continuous(breaks = seq(0, 5, by = 1)) +
  labs(x = "Date",
       y = "Total  MODIS GPP observations") +
  theme_light(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, h = 1))
```

### GPP, NDVI, and EVI seasonal patterns

Given that the time range of the data set selected for this research goes from
November 2012 to March 2017 and that Santa Rosa National Park presents a rainy
season and a dry season, I want to check first if there those seasonal patterns
are reflected in the MODIS products:

```{r}
modis_gpp_clean %>% 
  ggplot(aes(x = date, y = gpp, group = 1)) +
  geom_line(linetype = 2) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "GPP (mg m-2 s-1)")

modis_indices_clean %>% 
  mutate(date = ymd(date)) %>% 
  filter(ndvi > 0.2) %>% 
  ggplot(aes(x = date, y = ndvi)) +
  geom_jitter(alpha = 0.7, color = "#FF3A1D", size = 3) +
  geom_line(linetype = 2) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "NDVI")
```

### What is the relation between EVI NDVI & GPP MODIS products?

```{r MODIS data quality filtering}
## Filter observations with good quality
## This takes from 94 observations to 53 observations. Remove 41 observations
modis_indices_clean <- modis_indices %>% 
  filter(reliability_modland_description == "Good data, use with confidence") %>% 
  select(date, ndvi, evi) 

## Filter observations with good quality
## This takes from 183 observations to 114. Remove 69 observations.
modis_gpp_clean <- modis_gpp %>% 
  filter(modland_description == "Good quality",
         cloud_state_description == "Significant clouds NOT present (clear)",
         scf_qc_description == "Very best possible") %>% 
  select(date, gpp) 

modis_join <- modis_gpp_clean %>% 
  full_join(modis_indices_clean, by = c("date"))
```

We are using products from MODIS which are: EVI, NDVI and GPP. These are already
calculated values that are available to users. These products comes with
variables that flags low quality values. In this case we have a data set with
flags for GPP and a data set with flags for the indices EVI and NDVI.

All values with flags that advertised low quality data points were remove from
both data sets. Then a pearson correlation was performed to explore the relation
between GPP and NDVI, and GPP with EVI.

As validation for the relations between products from MODIS, I proceed to
compare NDVI and EVI against GPP to check how is the relation between these
variables.

```{r, fig.cap = "Relation between NDVI and GPP products from MODIS for the Santa Rosa National Park"}
#### Relation between NDVI and GPP from MODIS
gpp_ndvi <- modis_join %>% 
  ggplot(aes(x = ndvi, y = gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  # theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "NDVI", y = "GPP (mgm-2s-1)")

#### Relation between EVI and GPP from MODIS
gpp_evi <- modis_join %>% 
  ggplot(aes(x = evi, y = gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  # theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "EVI", y = "GPP (mg m-2 s-1)")

## Arrange both plots in one figure:
plot_grid(gpp_ndvi, gpp_evi, labels = c('A', 'B'), label_size = 12)
```

```{r corr results MODIS products}
ndvi_corr_test <- cor.test(modis_join$ndvi, modis_join$gpp)
evi_corr_test <- cor.test(modis_join$evi, modis_join$gpp)

# Check the report of the result
# report(ndvi_corr_test)
# report(evi_corr_test)
```

The Pearson's product-moment correlation between **NDVI and GPP** products from
MODIS is positive and statistically significant. (r = 0.81, 95% CI [0.65, 0.90],
t(34) = 8.01, p \< .001). For **EVI and GPP** the Pearson's product-moment
correlation is is positive, and statistically significant (r = 0.70, 95% CI
[0.48, 0.84], t(34) = 5.70, p \< .001)

### MODIS NDVI, EVI & GPP trends over time

```{r fig.cap = "GPP trends over the years for MODIS and in-situ data. Despite having similar trends, range is higher for in-situ data than the MODIS derived GPP", layout = "l-body-outset"}
# Plot with both GPP's
rec_monthly_gpp <- monthly_gpp %>% 
  select(date, average_gpp) %>% 
  rename(gpp = average_gpp) %>% 
  mutate(origin = "in_situ")

rec_modis_gpp <- modis_gpp %>% 
  select(date, gpp) %>% 
  mutate(date = floor_date(date, "month")) %>% 
  group_by(date) %>% 
  summarise(
    gpp = mean(gpp, na.rm = TRUE)
  ) %>% 
  mutate(origin = "modis") %>% 
  filter(date > ymd("2013-04-30"))

rec_both <- bind_rows(rec_monthly_gpp, rec_modis_gpp)

## Correlation trends between both GPP's`
rec_both %>% 
  ggplot(aes(x = date, y = gpp, group = origin)) +
  geom_line(aes(color = origin), size = 1) +
  theme_light(base_size = 10) +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "GPP (mg m-2 s-1)", color = "GPP source")
```

## Summary

This is a bullet list with the main insights from the exploratory data analysis
that can help to understand the models and the phenomena explained in the
[Results and Discussion
section](https://ronnyhdez.github.io/gpp_remote_sensing/results.html)

 - NDVI have a higher positive correlation with GPP product from MODIS than
   EVI.
 - Year 2015 was the lowest year in precipitation
 - GPP estimated in-situ does not match the pattern of GPP obtained from the
 MODIS product.
    
# References

> Castro, S. M., Sanchez-Azofeifa, G. A., & Sato, H. (2018). Effect of drought
> on productivity in a Costa Rican tropical dry forest. Environmental Research
> Letters, 13(4), 045001.

> Sánchez‐Azofeifa, G. A., Quesada, M., Rodríguez, J. P., Nassar, J. M., Stoner, K. E., Castillo, A., ... & Cuevas‐Reyes, P. (2005). Research priorities for Neotropical dry forests 1. Biotropica: The Journal of Biology and Conservation, 37(4), 477-485.
