---
title: "Methods"
description: |
  Welcome to the website. I hope you enjoy it!
site: distill::distill_website
output:
  distill::distill_article:
    code_folding: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(leaflet)
```

## Methods


3. Methods (~200 words)

Data collection – Briefly describe your sampling procedure/experimental 
design/measurement protocols. How did you select a sample from a population in
a way that allows you to generalize? Please only describe the data that you 
actually analyze as part of this course project, not other things you have done 
as part of your thesis!

Visual Elements – These are probably most important in this section to keep your
reader’s attention. Avoid lengthy method descriptions, and instead use maps of 
your study sites, diagrams of your experimental and sampling designs, and tables
with design statistics, such as number of sites/blocks/reps, and transects/plots
within reps. Ground or aerial photos of your field sites with diagrams drawn on 
top of them in PowerPoint can be very effective.

## Site

```{r}
sr_map <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  setView(lng = -85.6288, lat = 10.8529, zoom = 10) %>% 
  addMarkers(lng = -85.6288, lat = 10.8529, 
             popup = "Carbon Flux Tower at Santa Rosa National Park")

sr_map
```

## References

Didan, K. (2015). MOD13Q1 MODIS/Terra Vegetation Indices 16-Day L3 Global 250m SIN Grid V006. NASA EOSDIS Land Processes DAAC. Accessed 2022-02-28 from https://doi.org/10.5067/MODIS/MOD13Q1.006. Accessed February 28, 2022.

Running, S., Mu, Q., Zhao, M. (2015). MYD17A2H MODIS/Aqua Gross Primary Productivity 8-Day L4 Global 500m SIN Grid V006. NASA EOSDIS Land Processes DAAC. Accessed 2022-02-28 from https://doi.org/10.5067/MODIS/MYD17A2H.006. Accessed February 28, 2022.

AppEEARS Team. (2022). Application for Extracting and Exploring Analysis Ready Samples (AppEEARS). Ver. 2.71. NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS/Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, USA. Accessed February 28, 2022. https://lpdaacsvc.cr.usgs.gov/appeears
