---
title: "About this site"
description: |
  Some additional details about the website
site: distill::distill_website
output:
  distill::distill_article:
    code_folding: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(visdat)
library(corrplot)
library(tidyr)
library(rmarkdown)


precipitation <- 
  read_csv("data/hardvard_dataverse_tropidry/Daily_Precipitation_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd(date),
         do_y = as.numeric(do_y),
         precip = as.numeric(precip))

biomet <- 
  read_csv("data/hardvard_dataverse_tropidry/Micrometeorological_Biomet_Parameters_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd_hm(date)) %>% 
  mutate(across(where(is.character), as.numeric))

monthly_gpp <- 
  read_csv("data/hardvard_dataverse_tropidry/Monthly_NEE_GPP_Reco_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  unite(col = date, year, month, sep = "-") %>% 
  mutate(date = paste0(date, "-1"),
         date = ymd(date)) %>% 
  select(-date_1, -date_2) %>% 
  mutate(across(where(is.character), as.numeric))

ndvi_moha <- 
  readxl::read_xlsx("data/hardvard_dataverse_tropidry/final -NDVI- smoothed -complete.xlsx") %>% 
  clean_names()

modis_gpp <- 
  read_csv("data/modis-gpp-sr/modis-sr-MYD17A2H-006-results.csv") %>% 
  clean_names() %>% 
  select(date, myd17a2h_006_gpp_500m, 
         myd17a2h_006_psn_qc_500m_modland_description,
         myd17a2h_006_psn_qc_500m_cloud_state_description,
         myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  rename(gpp = myd17a2h_006_gpp_500m,
         modland_description = myd17a2h_006_psn_qc_500m_modland_description,
         cloud_state_description = myd17a2h_006_psn_qc_500m_cloud_state_description,
         scf_qc_description = myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  mutate(gpp = gpp * 10)

modis_indices <- 
  read_csv("data/modis-indices-sr/modis-indices-sr-MOD13Q1-006-results.csv") %>% 
  clean_names() %>% 
  select(date, mod13q1_006_250m_16_days_evi, 
         mod13q1_006_250m_16_days_ndvi, 
         mod13q1_006_250m_16_days_vi_quality,
         mod13q1_006_250m_16_days_pixel_reliability,
         mod13q1_006_250m_16_days_pixel_reliability_modland_description) %>% 
  rename(evi = mod13q1_006_250m_16_days_evi,
         ndvi = mod13q1_006_250m_16_days_ndvi,
         vi_quality = mod13q1_006_250m_16_days_vi_quality,
         pixel_reliability = mod13q1_006_250m_16_days_pixel_reliability,
         reliability_modland_description =
           mod13q1_006_250m_16_days_pixel_reliability_modland_description)
```

5. Results (including Discussion ~500 words)

Results and Discussion – Be selective and only show a reasonable number of 
quality graphs that describe your results. Discuss each graph (or group of 
graphs) with a separate paragraph that makes references to the figures (or 
tables) that you are talking about. Tell people what they see in the graph, 
point out interesting relationships, explain how they can be biologically 
interpreted, and/or what the practical applications of these findings are.

[For the draft submission, you are somewhat limited to complete this section.
You may apply what you learned in Labs 1 through 5 where applicable and useful.
Leave the rest for completion for the final submission. If you have some ideas 
which advanced multivariate methods you want to use here, you can briefly 
describe your plans to receive some feedback on that.]

Conclusions, About, References – You may add these as additional pages, 
paragraphs or footnotes. 


```{r}
# Plot with both GPP's
rec_monthly_gpp <- monthly_gpp %>% 
  select(date, average_gpp) %>% 
  rename(gpp = average_gpp) %>% 
  mutate(origin = "in_situ")

rec_modis_gpp <- modis_gpp %>% 
  select(date, gpp) %>% 
  mutate(date = floor_date(date, "month")) %>% 
  group_by(date) %>% 
  summarise(
    gpp = mean(gpp, na.rm = TRUE)
  ) %>% 
  mutate(origin = "modis")

rec_both <- bind_rows(rec_monthly_gpp, rec_modis_gpp)

rec_both %>% 
  ggplot(aes(x = date, y = gpp, group = origin)) +
  geom_line(aes(color = origin)) +
  theme_linedraw() +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "Date", y = "GPP (mgm-2s-1)", color = "Origin of GPP values")
```

