---
title: "Results and conclussions"
description: |
  MODIS products can understimate GPP in Santa Rosa National Park Tropical Dry Forest
site: distill::distill_website
output:
  distill::distill_article:
    code_folding: true
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Libraries
library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(visdat)
library(corrplot)
library(tidyr)
library(rmarkdown)
library(tidymodels)
library(report)
library(cowplot)
library(ecodist)
library(vegan)

# Import data sets
precipitation <- 
  read_csv("data/hardvard_dataverse_tropidry/Daily_Precipitation_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd(date),
         do_y = as.numeric(do_y),
         precip = as.numeric(precip))

biomet <- 
  read_csv("data/hardvard_dataverse_tropidry/Micrometeorological_Biomet_Parameters_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd_hm(date)) %>% 
  mutate(across(where(is.character), as.numeric))

monthly_gpp <- 
  read_csv("data/hardvard_dataverse_tropidry/Monthly_NEE_GPP_Reco_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  unite(col = date, year, month, sep = "-") %>% 
  mutate(date = paste0(date, "-1"),
         date = ymd(date)) %>% 
  select(-date_1, -date_2) %>% 
  mutate(across(where(is.character), as.numeric))

ndvi_moha <- 
  readxl::read_xlsx("data/hardvard_dataverse_tropidry/final -NDVI- smoothed -complete.xlsx") %>% 
  clean_names()

modis_gpp <- 
  read_csv("data/modis-gpp-sr/modis-sr-MYD17A2H-006-results.csv") %>% 
  clean_names() %>% 
  select(date, myd17a2h_006_gpp_500m, 
         myd17a2h_006_psn_qc_500m_modland_description,
         myd17a2h_006_psn_qc_500m_cloud_state_description,
         myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  rename(gpp = myd17a2h_006_gpp_500m,
         modland_description = myd17a2h_006_psn_qc_500m_modland_description,
         cloud_state_description = myd17a2h_006_psn_qc_500m_cloud_state_description,
         scf_qc_description = myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  mutate(gpp = gpp * 10)

modis_indices <- 
  read_csv("data/modis-indices-sr/modis-indices-sr-MOD13Q1-006-results.csv") %>% 
  clean_names() %>% 
  select(date, mod13q1_006_250m_16_days_evi, 
         mod13q1_006_250m_16_days_ndvi, 
         mod13q1_006_250m_16_days_vi_quality,
         mod13q1_006_250m_16_days_pixel_reliability,
         mod13q1_006_250m_16_days_pixel_reliability_modland_description) %>% 
  rename(evi = mod13q1_006_250m_16_days_evi,
         ndvi = mod13q1_006_250m_16_days_ndvi,
         vi_quality = mod13q1_006_250m_16_days_vi_quality,
         pixel_reliability = mod13q1_006_250m_16_days_pixel_reliability,
         reliability_modland_description =
           mod13q1_006_250m_16_days_pixel_reliability_modland_description)

## Filter observations with good quality
## This takes from 94 observations to 53 observations. Remove 41 observations
modis_indices_clean <- modis_indices %>% 
  filter(reliability_modland_description == "Good data, use with confidence") %>% 
  select(date, ndvi, evi) #%>% 
  # mutate(week = week(date),
  #        year = year(date))

## Filter observations with good quality
## This takes from 183 observations to 114. Remove 69 observations.
modis_gpp_clean <- modis_gpp %>% 
  filter(modland_description == "Good quality",
         cloud_state_description == "Significant clouds NOT present (clear)",
         scf_qc_description == "Very best possible") %>% 
  select(date, gpp) #%>% 
  # mutate(week = week(date),
  #        year = year(date))

modis_join <- modis_gpp_clean %>% 
  full_join(modis_indices_clean, by = c("date"))
```

<!-- 5. Results (including Discussion ~500 words) -->

<!-- Results and Discussion – Be selective and only show a reasonable number of  -->

<!-- quality graphs that describe your results. Discuss each graph (or group of  -->

<!-- graphs) with a separate paragraph that makes references to the figures (or  -->

<!-- tables) that you are talking about. Tell people what they see in the graph,  -->

<!-- point out interesting relationships, explain how they can be biologically  -->

<!-- interpreted, and/or what the practical applications of these findings are. -->

<!-- [For the draft submission, you are somewhat limited to complete this section. -->

<!-- You may apply what you learned in Labs 1 through 5 where applicable and useful. -->

<!-- Leave the rest for completion for the final submission. If you have some ideas  -->

<!-- which advanced multivariate methods you want to use here, you can briefly  -->

<!-- describe your plans to receive some feedback on that.] -->

<!-- Conclusions, About, References – You may add these as additional pages,  -->

<!-- paragraphs or footnotes.  -->

## Results

### What is the relation between MODIS products and GPP estimated in-situ?

We have two indices derived from MODIS that can have a relation with the
estimation of GPP in-situ. A linear model was performed to evaluate the 
relation between NDVI with GPP and, EVI with GPP estimated in-situ.

The clean data sets from MODIS products were used here altogether with the GPP
data set with the estimations per month from Santa Rosa National Park.

```{r corr_gpp_modis_in_situ, include = FALSE}
average_gpp <- modis_gpp_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    average_modis_gpp = mean(gpp)
  ) %>% 
  rename(month_year = `zoo::as.yearmon(date)`)

modis_in_situ_gpp <- monthly_gpp %>% 
  mutate(month_year = zoo::as.yearmon(date)) %>% 
  select(month_year, average_gpp) %>% 
  full_join(average_gpp, by = "month_year")

# modis_in_situ_gpp %>% 
#   ggplot(aes(x = average_gpp, y = average_modis_gpp)) +
#   geom_point(color = "#FF3A1D", size = 3.5) +
#   geom_smooth(method = "lm", color = "#4E5C68") +
#   theme_light(base_size = 12) +
#   labs(x = "GPP (mg m-2 s-1)", y = "GPP (mg m-2 s-1)")

# gpp_modis_in_situ <- cor.test(modis_in_situ_gpp$average_gpp,
#                               modis_in_situ_gpp$average_modis_gpp)
```

#### MODIS NDVI and GPP in-situ

Given that observations for GPP in-situ are one per month and we have in most
cases for the same period of time, more than one observation from MODIS, I have
summarized MODIS values with the mean per month. A linear regression was
performed and residuals were evaluated to look for patterns.

```{r linear_models}
## Prepare data for model
ndvi_gpp_in_situ <- modis_indices_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    ndvi_mean = mean(ndvi, na.rm = TRUE),
    evi_mean = mean(evi, na.rm = TRUE)
  ) %>% 
  rename(month_year = `zoo::as.yearmon(date)`) %>%
  full_join(modis_in_situ_gpp, by = "month_year")

## Create the linear models and check results
ndvi_gpp_model <- lm(average_gpp ~ ndvi_mean, data = ndvi_gpp_in_situ)
evi_gpp_model <- lm(average_gpp ~ evi_mean, data = ndvi_gpp_in_situ)

## Report models results
# summary(ndvi_gpp_model)
# report(ndvi_gpp_model)
# plot(evi_gpp_model)
# summary(evi_gpp_model)
# report(evi_gpp_model)
```

```{r fig.cap = "Relation between NDVI and EVI indices derived from MODIS with GPP estimated in-situ"}
ndvi_gpp <- ndvi_gpp_in_situ %>% 
  ggplot(aes(x = ndvi_mean, y = average_gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  labs(x = "MODIS NDVI", y = "GPP (mg m-2 s-1)")

evi_gpp <- ndvi_gpp_in_situ %>% 
  ggplot(aes(x = evi_mean, y = average_gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  labs(x = "MODIS EVI", y = "GPP (mg m-2 s-1)")

## Arrange both plots in one figure:
plot_grid(ndvi_gpp, evi_gpp, labels = c('A', 'B'), label_size = 12)
```

I fitted a linear model (estimated using OLS) to predict GPP in-situ with MODIS
NDVI product (formula: GPP in-situ \~ NDVI) as shown in Figure 1.A. The model 
explains a statistically significant and substantial proportion of variance 
(R2 = 0.84, F(1, 28) = 152.26, p \< .001, adj. R2 = 0.84). The model's intercept, corresponding to NDVI = 0, is at -0.70 (95% CI [-0.89, -0.50], t(28) = -7.35,
p \< .001). Within this model:

-   The effect of NDVI is statistically significant and positive (beta = 1.66,
    95% CI [1.38, 1.94], t(28) = 12.34, p \< .001; Std. beta = 0.92, 95% CI
    [0.77, 1.07])

A second linear model was fitted (estimated using OLS) to predict in-situ GPP 
with MODIS EVI product (formula: GPP in-situ ~ EVI) as shown in Figure 1.B. The 
model explains a statistically significant and substantial proportion of variance 
(R2 = 0.79, F(1, 28) = 102.72, p < .001, adj. R2 = 0.78). The model's intercept, 
corresponding to EVI = 0, isat -0.40 (95% CI [-0.58, -0.22], t(28) = -4.61, 
p < .001). Within this model:

  - The effect of EVI is statistically significant and positive (beta = 1.95, 
  95% CI [1.56, 2.35], t(28) = 10.14, p < .001; Std. beta = 0.89, 
  95% CI [0.71, 1.07])

For both models, standardized parameters were obtained by fitting the model on 
a standardized version of the dataset. 95% Confidence Intervals (CIs) and 
p-values were computed using the Wald approximation.

##### Residuals exploration

The residuals from both linear models were used to explore any further patterns
in time. Specifically, if there are months in which residuals are larger or
smaller consistently along the time period of this study.

```{r fig.cap = "Residuals values for linear models GPP ~ NDVI (A) and, GPP ~ EVI (B) for registered months", layout="l-body-outset"}
## GPP ~ NDVI residuals %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
## Obtain data used in the model (without NA's) to paste residuals and explore
## if there is a pattern of largest residuals in specific months
model_observations <- ndvi_gpp_model[["model"]]
model_residuals <- ndvi_gpp_model[["residuals"]] %>% 
  as.data.frame() %>% 
  rename("residuals" = ".")

# Create dataset
ndvi_observations_residuals <- model_observations %>% 
  left_join(ndvi_gpp_in_situ, by = c("average_gpp", "ndvi_mean")) %>% 
  select(-average_modis_gpp, -evi_mean) %>% 
  bind_cols(model_residuals)

## Plot with distances from zero:
ndvi_residuals <- ndvi_observations_residuals %>% 
  mutate(residuals = round(residuals, digits = 3)) %>% 
  mutate(date = zoo::as.Date(month_year)) %>% 
  ggplot(aes(x = date, y = residuals, label = residuals)) + 
  geom_point(stat = 'identity', color = "#75AADB", size = 9)  +
  geom_hline(yintercept = 0, color = "#4E5C68",
             size = 0.5, linetype = "dashed") +
  scale_x_date(date_labels = "%b-%Y",  date_breaks = "2 month") +
  geom_segment(aes(y = 0, 
                   x = date, 
                   yend = residuals, 
                   xend = date), 
               color = "#75AADB") +
  geom_text(color = "white", size = 2) +
  labs(y = "Residuals GPP ~ MODIS_NDVI",
       x = "Date") + 
  ylim(-0.3, 0.3) +
  theme_light(base_size = 10) +
  coord_flip()

## GPP ~ EVI residuals %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
## Obtain data used in the model (without NA's) to paste residuals and explore
## if there is a pattern of largest residuals in specific months
model_observations <- evi_gpp_model[["model"]]
model_residuals <- evi_gpp_model[["residuals"]] %>% 
  as.data.frame() %>% 
  rename("residuals" = ".")

# Create dataset
evi_observations_residuals <- model_observations %>% 
  left_join(ndvi_gpp_in_situ, by = c("average_gpp", "evi_mean")) %>% 
  select(-average_modis_gpp, -evi_mean) %>% 
  bind_cols(model_residuals)

## Plot with distances from zero:
evi_residuals <- evi_observations_residuals %>% 
  mutate(residuals = round(residuals, digits = 3)) %>% 
  mutate(date = zoo::as.Date(month_year)) %>% 
  ggplot(aes(x = date, y = residuals, label = residuals)) + 
  geom_point(stat = 'identity', color = "#75AADB", size = 9)  +
  geom_hline(yintercept = 0, color = "#4E5C68",
             size = 0.5, linetype = "dashed") +
  scale_x_date(date_labels = "%b-%Y",  date_breaks = "2 month") +
  geom_segment(aes(y = 0, 
                   x = date, 
                   yend = residuals, 
                   xend = date), 
               color = "#75AADB") +
  geom_text(color = "white", size = 2) +
  labs(y = "Residuals GPP ~ MODIS_EVI",
       x = "Date") + 
  ylim(-0.3, 0.3) +
  theme_light(base_size = 10) +
  coord_flip() 

## Arrange both plots in one figure:
plot_grid(ndvi_residuals, evi_residuals, labels = c('A', 'B'), label_size = 12)
```

Figure 2 shows the residuals values derived from the linear model between
GPP with NDVI (A), and GPP with EVI (B). No patterns or consistency between
months can be inferred from Figure 2.

##### Residuals and meteorological variables

Given that there is no clear pattern in the values of the residuals with just
the months and years, I made a comparison between the values of the residuals
with meteorological values per month.

 - Do we have meteorological characteristics that are related to larger or 
 shorter residuals values?

```{r fig.cap = "Correlation between the NDVI and EVI residuals with meteorological variables"}
# NDVI BIOMET %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
observations_ndvi_residuals <- ndvi_observations_residuals %>% 
  select(month_year, residuals) %>% 
  rename(ndvi_residuals = residuals)

# EVI BIOMET %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
observations_evi_residuals <- evi_observations_residuals %>% 
  select(month_year, residuals) %>% 
  rename(evi_residuals = residuals)

# Summarise biomet %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
biomet_summarized <- biomet %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    par_incoming = mean(par_incoming, na.rm = TRUE),
    swc = mean(swc, na.rm = TRUE),
    vpd = mean(vpd, na.rm = TRUE),
    r_h = mean(r_h, na.rm = TRUE),
    tair = mean(tair, na.rm = TRUE),
    le = mean(le, na.rm = TRUE),
    h = mean(h, na.rm = TRUE)
  ) %>% 
  rename(month_year = `zoo::as.yearmon(date)`)

precipitation_summarized <- precipitation %>% 
  mutate(month_year = zoo::as.yearmon(date)) %>% 
  group_by(month_year) %>% 
  summarize(
    total_precip = sum(precip, na.rm = TRUE)
  )

# Join datasets %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
residuals_biomet <- observations_ndvi_residuals %>% 
  left_join(observations_evi_residuals, by = "month_year") %>% 
  left_join(biomet_summarized, by = "month_year") %>% 
  left_join(precipitation_summarized, by = "month_year") %>% 
  select(-month_year) 

tmwr_cols <- colorRampPalette(c("#91CBD765", "#CA225E"))

residuals_biomet %>% 
  drop_na() %>% 
  cor() %>% 
  corrplot(col = tmwr_cols(200), tl.col = "black")
```

 - There is no a clear correlation between residuals (from NDVI and EVI) with
 the meteorological variables, other than Latent Heat (`le`), and total 
 precipitation per month (`total_precip`)

```{r fig.cap = "Check PCA"}
# # For EVI
# biomet_pca <- residuals_biomet %>%
#   select(-ndvi_residuals) %>%
#   drop_na() %>%
#   princomp(cor = T)
# 
# biplot(biomet_pca)
# 
# # For NDVI
# biomet_pca <- residuals_biomet %>%
#   select(-evi_residuals) %>%
#   drop_na() %>%
#   princomp(cor = T)
# 
# biplot(biomet_pca)
```

### Gradient analysis

In order to know which meteorological variables have more influences in the
values of GPP (in-situ and MODIS) and indices, a gradient analysis (direct and
indirect) was performed. Mean values for each month were calculated and those
months which had no value in one of the variables were filtered out, resulting
in a total of 26 observations (available months with complete values) for all
the time period in this research.

```{r results = FALSE}
# biomet_summarized contains the meteorological variables per month
# modis_indices_clean
# modis_gpp_clean

# Unite MODIS products per month and year %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
modis_gpp_month <- modis_gpp_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    gpp = mean(gpp, na.rm = TRUE)
  ) %>% 
  rename(date = `zoo::as.yearmon(date)`)

modis_indices_month <- modis_indices_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    ndvi = mean(ndvi, na.rm = TRUE),
    evi = mean(evi, na.rm = TRUE)
  ) %>% 
  rename(date = `zoo::as.yearmon(date)`)

modis_means <- modis_gpp_month %>% 
  full_join(modis_indices_month, by = "date")

# Unite MODIS products with gpp in situ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
in_situ_gpp <- monthly_gpp %>% 
  mutate(date = zoo::as.yearmon(date)) %>% 
  select(date, average_gpp) %>% 
  rename(gpp_in_situ = average_gpp)

full_gpp <- in_situ_gpp %>% 
  full_join(modis_means) %>% 
  rename(gpp_modis = gpp,
         month_year = date) %>% 
  full_join(biomet_summarized) %>% 
  drop_na()

# Prepare data for gradient analysis %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
rownames(full_gpp) = full_gpp$month_year
month_labels = as.factor(full_gpp$month_year)

indices <- full_gpp[ , 2:5]
meteo <- full_gpp[ , 6:12]

# NDMS execution %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
sed <-  distance(scale(meteo), "euclidean")
out1 <-  metaMDS(sed, k = 2, trymax = 500)
scores1 <-  out1$points # generate scores
```

#### Direct gradient analysis

```{r fig.cap = "Direct gradient analysis", layout="l-body-outset", fig.height = 7}
# Plot ordination variables %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
plot(scores1, col = "white")
text(scores1, labels = month_labels)
vectors1 = envfit(scores1, meteo, nperm = 0)
plot(vectors1, col = "red")

# Adding new vectors %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
vectors2 = envfit(scores1, indices, nperm = 999)
plot(vectors2, col = "blue")
```

```{r, results = FALSE}
dm2 <- distance(indices, "euclidean")
out2 <- metaMDS(dm2, k = 2, trymax = 500)
scores2 <- out2$points # generate scores
```

The scores for this analysis are:

```{r}
vectors2
```

#### Indirect gradient analysis

```{r fig.cap = "Indirect gradient analysis", layout="l-body-outset", fig.height = 7}
plot(scores2, col = "white")
text(scores2, labels = month_labels)
vectors3 <- envfit(scores2, indices, nperm = 0)
plot(vectors3, len = 0.1, col = "red")
vectors4 <- envfit(scores2, meteo, nperm = 999)
plot(vectors4, len = 0.1, col = "blue")
```

The scores for this analysis are:

```{r}
vectors4
```

## Conclusions

 - Given data constrains (4 years of data from a single point) no further 
 analysis could be done. Nonetheless, climate variables relations with driest
 months can indicate an influence on the estimations of GPP through MODIS
 products
 - MODIS NDVI product have a better relation with GPP estimated in-situ than EVI
 MODIS product
 - GPP derived from MODIS does not follow the seasonal patterns of GPP estimated
 in-situ.
 - Soil water content, relative humidity and latent heat are more associated with
 months with higher precipitation and MODIS products.


## References

> James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013). An introduction to
> statistical learning (Vol. 112, p. 18). New York: springer.
