---
title: "Results and conclussions"
description: |
  MODIS products can understimate GPP in Santa Rosa National Park Tropical Dry Forest
site: distill::distill_website
output:
  distill::distill_article:
    code_folding: true
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Libraries
library(readr)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(visdat)
library(corrplot)
library(tidyr)
library(rmarkdown)
library(tidymodels)
library(report)
library(cowplot)

# Import data sets
precipitation <- 
  read_csv("data/hardvard_dataverse_tropidry/Daily_Precipitation_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd(date),
         do_y = as.numeric(do_y),
         precip = as.numeric(precip))

biomet <- 
  read_csv("data/hardvard_dataverse_tropidry/Micrometeorological_Biomet_Parameters_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  mutate(date = ymd_hm(date)) %>% 
  mutate(across(where(is.character), as.numeric))

monthly_gpp <- 
  read_csv("data/hardvard_dataverse_tropidry/Monthly_NEE_GPP_Reco_2013_to_2016.csv") %>% 
  clean_names() %>% 
  slice(-1) %>% 
  unite(col = date, year, month, sep = "-") %>% 
  mutate(date = paste0(date, "-1"),
         date = ymd(date)) %>% 
  select(-date_1, -date_2) %>% 
  mutate(across(where(is.character), as.numeric))

ndvi_moha <- 
  readxl::read_xlsx("data/hardvard_dataverse_tropidry/final -NDVI- smoothed -complete.xlsx") %>% 
  clean_names()

modis_gpp <- 
  read_csv("data/modis-gpp-sr/modis-sr-MYD17A2H-006-results.csv") %>% 
  clean_names() %>% 
  select(date, myd17a2h_006_gpp_500m, 
         myd17a2h_006_psn_qc_500m_modland_description,
         myd17a2h_006_psn_qc_500m_cloud_state_description,
         myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  rename(gpp = myd17a2h_006_gpp_500m,
         modland_description = myd17a2h_006_psn_qc_500m_modland_description,
         cloud_state_description = myd17a2h_006_psn_qc_500m_cloud_state_description,
         scf_qc_description = myd17a2h_006_psn_qc_500m_scf_qc_description) %>% 
  mutate(gpp = gpp * 10)

modis_indices <- 
  read_csv("data/modis-indices-sr/modis-indices-sr-MOD13Q1-006-results.csv") %>% 
  clean_names() %>% 
  select(date, mod13q1_006_250m_16_days_evi, 
         mod13q1_006_250m_16_days_ndvi, 
         mod13q1_006_250m_16_days_vi_quality,
         mod13q1_006_250m_16_days_pixel_reliability,
         mod13q1_006_250m_16_days_pixel_reliability_modland_description) %>% 
  rename(evi = mod13q1_006_250m_16_days_evi,
         ndvi = mod13q1_006_250m_16_days_ndvi,
         vi_quality = mod13q1_006_250m_16_days_vi_quality,
         pixel_reliability = mod13q1_006_250m_16_days_pixel_reliability,
         reliability_modland_description =
           mod13q1_006_250m_16_days_pixel_reliability_modland_description)
```

<!-- 5. Results (including Discussion ~500 words) -->

<!-- Results and Discussion – Be selective and only show a reasonable number of  -->

<!-- quality graphs that describe your results. Discuss each graph (or group of  -->

<!-- graphs) with a separate paragraph that makes references to the figures (or  -->

<!-- tables) that you are talking about. Tell people what they see in the graph,  -->

<!-- point out interesting relationships, explain how they can be biologically  -->

<!-- interpreted, and/or what the practical applications of these findings are. -->

<!-- [For the draft submission, you are somewhat limited to complete this section. -->

<!-- You may apply what you learned in Labs 1 through 5 where applicable and useful. -->

<!-- Leave the rest for completion for the final submission. If you have some ideas  -->

<!-- which advanced multivariate methods you want to use here, you can briefly  -->

<!-- describe your plans to receive some feedback on that.] -->

<!-- Conclusions, About, References – You may add these as additional pages,  -->

<!-- paragraphs or footnotes.  -->

## Results

### What is the relation between MODIS EVI NDVI & GPP products?

We are using products from MODIS which are: EVI, NDVI and GPP. These are already
calculated values that are available to users. These products comes with
variables that flags low quality values. In this case we have a data set with
flags for GPP and a data set with flags for the indices EVI and NDVI.

All values with flags that advertised low quality data points were remove from
both data sets. Then a pearson correlation was performed to explore the relation
between GPP and NDVI, and GPP with EVI.

```{r MODIS data quality filtering}
## Filter observations with good quality
## This takes from 94 observations to 53 observations. Remove 41 observations
modis_indices_clean <- modis_indices %>% 
  filter(reliability_modland_description == "Good data, use with confidence") %>% 
  select(date, ndvi, evi) #%>% 
  # mutate(week = week(date),
  #        year = year(date))

## Filter observations with good quality
## This takes from 183 observations to 114. Remove 69 observations.
modis_gpp_clean <- modis_gpp %>% 
  filter(modland_description == "Good quality",
         cloud_state_description == "Significant clouds NOT present (clear)",
         scf_qc_description == "Very best possible") %>% 
  select(date, gpp) #%>% 
  # mutate(week = week(date),
  #        year = year(date))

modis_join <- modis_gpp_clean %>% 
  full_join(modis_indices_clean, by = c("date"))
```

As validation for the relations between products from MODIS, I proceed to
compare NDVI against GPP to check how is the relation between these variables.

```{r, fig.cap = "Relation between NDVI and GPP products from MODIS for the Santa Rosa National Park"}
#### Relation between NDVI and GPP from MODIS
gpp_ndvi <- modis_join %>% 
  ggplot(aes(x = ndvi, y = gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  # theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "NDVI", y = "GPP (mgm-2s-1)")

#### Relation between EVI and GPP from MODIS
gpp_evi <- modis_join %>% 
  ggplot(aes(x = evi, y = gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  # theme(axis.text.x = element_text(angle = 90, h = 1)) +
  labs(x = "EVI", y = "GPP (mg m-2 s-1)")

## Arrange both plots in one figure:
plot_grid(gpp_ndvi, gpp_evi, labels = c('A', 'B'), label_size = 12)
```

```{r corr results MODIS products}
ndvi_corr_test <- cor.test(modis_join$ndvi, modis_join$gpp)
evi_corr_test <- cor.test(modis_join$evi, modis_join$gpp)

# Check the report of the result
# report(ndvi_corr_test)
# report(evi_corr_test)
```

The Pearson's product-moment correlation between **NDVI and GPP** products from
MODIS is positive and statistically significant. (r = 0.81, 95% CI [0.65, 0.90],
t(34) = 8.01, p \< .001). For **EVI and GPP** the Pearson's product-moment
correlation is is positive, and statistically significant (r = 0.70, 95% CI
[0.48, 0.84], t(34) = 5.70, p \< .001)

-   NDVI have a higher positive correlation with GPP product from MODIS than
    EVI.

### What is the relation between MODIS products and GPP estimated in-situ?

Now that I checked the relation between the MODIS products, I can compared and
evaluate the relation between the MODIS products with the GPP estimated in-situ.
The clean data sets from MODIS products were used here altogether with the GPP
data set with the estimations per month from Santa Rosa National Park.

```{r corr_gpp_modis_in_situ}
average_gpp <- modis_gpp_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    average_modis_gpp = mean(gpp)
  ) %>% 
  rename(month_year = `zoo::as.yearmon(date)`)

modis_in_situ_gpp <- monthly_gpp %>% 
  mutate(month_year = zoo::as.yearmon(date)) %>% 
  select(month_year, average_gpp) %>% 
  full_join(average_gpp, by = "month_year")

modis_in_situ_gpp %>% 
  ggplot(aes(x = average_gpp, y = average_modis_gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  labs(x = "GPP (mg m-2 s-1)", y = "GPP (mg m-2 s-1)")

gpp_modis_in_situ <- cor.test(modis_in_situ_gpp$average_gpp,
                              modis_in_situ_gpp$average_modis_gpp)
```

#### MODIS NDVI and GPP in-situ

##### Linear model
Given that observations for GPP in-situ are one per month and we have in most
cases for the same period of time, more than one observation from MODIS, I have
summarized MODIS values with the mean per month. A linear regression was
performed and residuals were evaluated to look for patterns.

```{r}
ndvi_gpp_in_situ <- modis_indices_clean %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    ndvi_mean = mean(ndvi, na.rm = TRUE),
    evi_mean = mean(evi, na.rm = TRUE)
  ) %>% 
  rename(month_year = `zoo::as.yearmon(date)`) %>%
  full_join(modis_in_situ_gpp, by = "month_year")

## Create the linear model and check results
ndvi_gpp_model <- lm(average_gpp ~ ndvi_mean, data = ndvi_gpp_in_situ)
# summary(ndvi_gpp_model)
# report(ndvi_gpp_model)
```

I fitted a linear model (estimated using OLS) to predict GPP in-situ with MODIS
NDVI product (formula: GPP in-situ \~ NDVI). The model explains a statistically
significant and substantial proportion of variance (R2 = 0.84, F(1, 28) =
152.26, p \< .001, adj. R2 = 0.84). The model's intercept, corresponding to NDVI
= 0, is at -0.70 (95% CI [-0.89, -0.50], t(28) = -7.35, p \< .001). Within this
model:

-   The effect of NDVI is statistically significant and positive (beta = 1.66,
    95% CI [1.38, 1.94], t(28) = 12.34, p \< .001; Std. beta = 0.92, 95% CI
    [0.77, 1.07])

Standardized parameters were obtained by fitting the model on a standardized
version of the data set. 95% Confidence Intervals (CIs) and p-values were
computed using the Wald approximation.

```{r}
ndvi_gpp_in_situ %>% 
  ggplot(aes(x = ndvi_mean, y = average_gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  labs(x = "MODIS NDVI", y = "GPP (mg m-2 s-1)")
```

##### Residuals

Where do we have the largest residuals? Is there a patter between the largest
and smallest residuals?

```{r}
## Obtain data used in the model (without NA's) to paste residuals and explore
## if there is a pattern of largest residuals in specific months
model_observations <- ndvi_gpp_model[["model"]]
model_residuals <- ndvi_gpp_model[["residuals"]] %>% 
  as.data.frame() %>% 
  rename("residuals" = ".")

# Create dataset
observations_residuals <- model_observations %>% 
  left_join(ndvi_gpp_in_situ, by = c("average_gpp", "ndvi_mean")) %>% 
  select(-average_modis_gpp, -evi_mean) %>% 
  bind_cols(model_residuals)

## Plot with distances from zero:
observations_residuals %>% 
  mutate(residuals = round(residuals, digits = 3)) %>% 
  mutate(date = zoo::as.Date(month_year)) %>% 
  ggplot(aes(x = date, y = residuals, label = residuals)) + 
  geom_point(stat = 'identity', color = "#75AADB", size = 9)  +
  geom_hline(yintercept = 0, color = "#4E5C68",
             size = 0.5, linetype = "dashed") +
  scale_x_date(date_labels = "%b-%Y",  date_breaks = "2 month") +
  geom_segment(aes(y = 0, 
                   x = date, 
                   yend = residuals, 
                   xend = date), 
               color = "#75AADB") +
  geom_text(color = "white", size = 2) +
  labs(y = "Residuals values linear relation GPP ~ MODIS_NDVI",
       x = "Date") + 
  ylim(-0.3, 0.3) +
  theme_light(base_size = 12) +
  coord_flip()
```

##### Residuals and meteorological variables

Now that we have which residuals are larger in which months, we can try to
validate if there is a common pattern in those months with the largest
residuals.

```{r}
## Table with months with higher residuals
## Cuadratic formula to eliminate negative values
observations_residuals <- observations_residuals %>% 
  mutate(positive_residuals = residuals ^ 2,
         positive_residuals = round(positive_residuals, digits = 3)) %>% 
  select(month_year, positive_residuals, residuals) %>% 
  arrange(desc(positive_residuals))
```

Now that we have the transformation of the residuals (all positive) I can try to
compare these values with the means of meteorological values per month to check
if there is a relation.

```{r}
## Take biomet variables to validate if there is a similar pattern:
biomet_summarized <- biomet %>% 
  group_by(zoo::as.yearmon(date)) %>% 
  summarise(
    par_incoming = mean(par_incoming, na.rm = TRUE),
    swc = mean(swc, na.rm = TRUE),
    vpd = mean(vpd, na.rm = TRUE),
    r_h = mean(r_h, na.rm = TRUE),
    tair = mean(tair, na.rm = TRUE),
    le = mean(le, na.rm = TRUE),
    h = mean(h, na.rm = TRUE)
  ) %>% 
  rename(month_year = `zoo::as.yearmon(date)`)

check <- observations_residuals %>% 
  left_join(biomet_summarized, by = "month_year") %>% 
  select(-month_year) 

# library(PerformanceAnalytics)
# chart.Correlation(check, histogram = TRUE, method = "pearson")
## Los residuales comparados parecen solo relacionarse con LE.
```

#### MODIS EVI and GPP in-situ?

##### Linear model

```{r}
# I already have ndvi_gpp_in_situ object

## Create the linear model and check results
evi_gpp_model <- lm(average_gpp ~ evi_mean, data = ndvi_gpp_in_situ)
# plot(evi_gpp_model)
# summary(evi_gpp_model)
# report(evi_gpp_model)
```

We fitted a linear model (estimated using OLS) to predict in-situ GPP with MODIS
EVI product (formula: GPP in-situ ~ EVI). The model explains a statistically significant and substantial proportion of variance (R2 = 0.79, F(1, 28) = 102.72,
p < .001, adj. R2 = 0.78). The model's intercept, corresponding to EVI = 0, is 
at -0.40 (95% CI [-0.58, -0.22], t(28) = -4.61, p < .001). Within this model:

  - The effect of EVI is statistically significant and positive (beta = 1.95, 
  95% CI [1.56, 2.35], t(28) = 10.14, p < .001; Std. beta = 0.89, 
  95% CI [0.71, 1.07])

Standardized parameters were obtained by fitting the model on a standardized 
version of the dataset. 95% Confidence Intervals (CIs) and p-values were
computed using the Wald approximation.

```{r}
ndvi_gpp_in_situ %>% 
  ggplot(aes(x = evi_mean, y = average_gpp)) +
  geom_point(color = "#FF3A1D", size = 3.5) +
  geom_smooth(method = "lm", color = "#4E5C68") +
  theme_light(base_size = 12) +
  labs(x = "MODIS NDVI", y = "GPP (mg m-2 s-1)")
```

##### Residuals 

Where do we have the largest residuals? Is there a patter between the largest
and smallest residuals?

```{r}
## Obtain data used in the model (without NA's) to paste residuals and explore
## if there is a pattern of largest residuals in specific months
model_observations <- evi_gpp_model[["model"]]
model_residuals <- evi_gpp_model[["residuals"]] %>% 
  as.data.frame() %>% 
  rename("residuals" = ".")

# Create dataset
observations_residuals <- model_observations %>% 
  left_join(ndvi_gpp_in_situ, by = c("average_gpp", "evi_mean")) %>% 
  select(-average_modis_gpp, -evi_mean) %>% 
  bind_cols(model_residuals)

## Plot with distances from zero:
observations_residuals %>% 
  mutate(residuals = round(residuals, digits = 3)) %>% 
  mutate(date = zoo::as.Date(month_year)) %>% 
  ggplot(aes(x = date, y = residuals, label = residuals)) + 
  geom_point(stat = 'identity', color = "#75AADB", size = 9)  +
  geom_hline(yintercept = 0, color = "#4E5C68",
             size = 0.5, linetype = "dashed") +
  scale_x_date(date_labels = "%b-%Y",  date_breaks = "2 month") +
  geom_segment(aes(y = 0, 
                   x = date, 
                   yend = residuals, 
                   xend = date), 
               color = "#75AADB") +
  geom_text(color = "white", size = 2) +
  labs(y = "Residuals values linear relation GPP ~ MODIS_EVI",
       x = "Date") + 
  ylim(-0.3, 0.3) +
  theme_light(base_size = 12) +
  coord_flip()
```

##### Residuals and meteorological variables

Now that we have which residuals are larger in which months, we can try to
validate if there is a common pattern in those months with the largest
residuals.

```{r}
## Table with months with higher residuals
## Cuadratic formula to eliminate negative values
observations_residuals <- observations_residuals %>% 
  mutate(positive_residuals = residuals ^ 2,
         positive_residuals = round(positive_residuals, digits = 3)) %>% 
  select(month_year, positive_residuals, residuals) %>% 
  arrange(desc(positive_residuals))
```

Now that we have the transformation of the residuals (all positive) I can try to
compare these values with the means of meteorological values per month to check
if there is a relation.

```{r}
## Take biomet variables to validate if there is a similar pattern:
# biomet_summarized is already a created object.
evi_biomet <- observations_residuals %>% 
  left_join(biomet_summarized, by = "month_year") %>% 
  select(-month_year) 

# library(PerformanceAnalytics)
# chart.Correlation(evi_biomet, histogram = TRUE, method = "pearson")
## Los residuales comparados parecen solo relacionarse con LE (igual que NDVI)
```

### Which meteorological variable have more influence in GPP in-situ?

```{r}

```

### MODIS NDVI, EVI & GPP trends over time

```{r fig.cap = "GPP trends over the years for MODIS and in-situ data. Despite having similar trends, range is higher for in-situ data than the MODIS derived GPP"}
# Plot with both GPP's
rec_monthly_gpp <- monthly_gpp %>% 
  select(date, average_gpp) %>% 
  rename(gpp = average_gpp) %>% 
  mutate(origin = "in_situ")

rec_modis_gpp <- modis_gpp %>% 
  select(date, gpp) %>% 
  mutate(date = floor_date(date, "month")) %>% 
  group_by(date) %>% 
  summarise(
    gpp = mean(gpp, na.rm = TRUE)
  ) %>% 
  mutate(origin = "modis") %>% 
  filter(date > ymd("2013-04-30"))

rec_both <- bind_rows(rec_monthly_gpp, rec_modis_gpp)

## Correlation trends between both GPP's`
rec_both %>% 
  ggplot(aes(x = date, y = gpp, group = origin)) +
  geom_line(aes(color = origin), size = 1) +
  theme_linedraw(base_size = 12) +
  scale_x_date(date_labels = "%b%Y", breaks = "months") +
  theme(axis.text.x = element_text(angle = 70, h = 1)) +
  labs(x = "Date", y = "GPP (mgm-2s-1)", color = "GPP source")
```

### PCA for meteorological variables

TODO: Implement nice PCA plot!!!

```{r}
# nrow(biomet)
# biomet %>% drop_na() %>% nrow()
biomet_pca <- biomet %>% 
  select(-date, -year, -do_y, -hour) %>% 
  drop_na() %>% 
  princomp(cor = T)

# biplot(biomet_pca)
```

### Gradient analysis

```{r}

```

#### Regression trees

This section is still a **Work In Progress**. The objective is to apply a
regression Random Forest using the meteorological variables as predictors of
GPP. After the evaluation of the predictions also a variable importance analysis
will be run.

Given that GPP observations are monthly, this reduces the total amount of
observations available, so bootstrapping is applied to the regression random
forest.

```{r}
ndvi_summarized <- ndvi_moha %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  group_by(year_mon) %>% 
  summarize(
    ndvi_mean = mean(ndvi, na.rm = TRUE),
    ndvi_sd = sd(ndvi, na.rm = TRUE)
  )

biomet_summarized <- biomet %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  group_by(year_mon) %>% 
  summarize(
    # sd_par_incoming = sd(par_incoming, na.rm = TRUE), 
    # sd_swc          = sd(swc, na.rm = TRUE), 
    # sd_vpd          = sd(vpd, na.rm = TRUE), 
    # sd_r_h          = sd(r_h, na.rm = TRUE), 
    # sd_tair         = sd(tair, na.rm = TRUE), 
    # sd_nee          = sd(nee, na.rm = TRUE), 
    # sd_le           = sd(le, na.rm = TRUE), 
    # sd_h            = sd(h, na.rm = TRUE),
    mean_par_incoming = sd(par_incoming, na.rm = TRUE), 
    mean_swc          = sd(swc, na.rm = TRUE), 
    mean_vpd          = sd(vpd, na.rm = TRUE), 
    mean_r_h          = sd(r_h, na.rm = TRUE), 
    mean_tair         = sd(tair, na.rm = TRUE), 
    # mean_nee          = sd(nee, na.rm = TRUE), 
    mean_le           = sd(le, na.rm = TRUE), 
    mean_h            = sd(h, na.rm = TRUE)
  )

summarize_dataset <- monthly_gpp %>% 
  mutate(year_mon = zoo::as.yearmon(date)) %>% 
  inner_join(ndvi_summarized) %>% 
  inner_join(biomet_summarized) %>% 
  drop_na()

set.seed(123)
gpp_split <- initial_split(summarize_dataset)
gpp_train <- training(gpp_split)
gpp_test <- testing(gpp_split)

set.seed(123)
gpp_boot <- bootstraps(gpp_train)
gpp_boot

rf_spec <- rand_forest() %>%
  set_mode("regression") %>%
  set_engine("ranger")

rf_spec

gpp_wf <- workflow() %>%
  add_formula(average_gpp ~ .)

gpp_wf


rf_rs <- gpp_wf %>%
  add_model(rf_spec) %>%
  fit_resamples(
    resamples = gpp_boot,
    control = control_resamples(save_pred = TRUE)
  )

rf_rs


## Evaluate model
collect_metrics(rf_rs)
```

## Conclusions

## References

> James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013). An introduction to
> statistical learning (Vol. 112, p. 18). New York: springer.
